---
title: "Causes of Splintering"
output: html_notebook
---

# Data Cleaning and Merging

In this analysis I'll be exploring the predictors of rebel group fragmentation. The first step is to import the UCDP actor data with my added variables, and clean it.

```{r data cleaning}
# Load and clean data
library(readr)
library(tidyr)

groups <- read_csv("rebel_origin.csv")
groups <- groups[c(1,35:43)]

#add in untouched version with correct encoding
groups2 <- read_csv("rebel_origin2.csv")
groups2 <- groups2[,1:34]
groups <- merge(groups2, groups, by = "actorid", all=T)
rm(groups2)

#remove cases that aren't single groups, cases w/ missing origin
groups <- subset(groups, is.na(delete)==T)
groups <- subset(groups, is.na(origin)==F)

groups <- subset(groups, select=-delete)

#split multiple origins
groups <- separate(groups, "origin",  c("origin","origin2"), sep="; ")

# Condense categories
groups$origin[groups$origin=="alliance - political parties"] <- "political party"
groups$origin[groups$origin=="criminal organization"] <- "economic organization"
groups$origin[groups$origin=="mercenaries"] <- "militia"
groups$origin[groups$origin=="protests"] <- "grassroots"
groups$origin[groups$origin=="anti-colonial"] <- "secessionist organization"
groups$origin[groups$origin=="underground resistance"] <- "political party"
groups$origin[groups$origin=="political organization"] <- "political party"
groups$origin[groups$origin=="transnational militant organization"] <- "transnational organization"
groups$origin[groups$origin=="transnational political organization"] <- "transnational organization"
groups$origin[groups$origin=="occupying military"] <- "military faction"

#separate ideology
groups <- separate(groups, "ideology", c("aim1", "aim2", "aim3"), sep="; ")

#write_csv(groups, "rebel_origins_clean.csv")
```

Then let's code the DV of whether a group splinters

```{r dv}

```

Next let's grab a few covariates based on the conflict data. The number of other groups present with it, the max and average intensity, the duration of time it is active.

```{r conflict}

```


Then I'll add the Gleditsch et al. data to get a bunch of covariates for most of my cases.

```{r gleditsch}
gled <- read_delim("gled.asc", delim="\t")


```

